---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

=======================================================


## Loading and preprocessing the data

Assumes data is in the user's working directory. Uses the data.table package to read the csv file into the object, "steps_data"

```{r}
library(data.table)
steps_data <- data.table(read.csv("activity.csv"))
```

Calculate the total number of steps for each day and produce a histogram.

```{r}
agg_steps <- as.data.frame(steps_data[,j=list(total_steps=sum(steps)),by=date])
hist(agg_steps$total_steps, xlab = "Total Steps", main = "Histogram of Daily Total Steps")
```

## What is mean total number of steps taken per day?

Caluculate mean and median values and display.

```{r averages}
mn <- round(mean(agg_steps$total_steps, na.rm=TRUE))
md <- median(agg_steps$total_steps, na.rm=TRUE)
```

* The mean number of total steps taken is `r mn`
* The median number of total steps taken is `r md`


## What is the average daily activity pattern?

Calculate the average steps by five minute interval and plot a line graph.

```{r}
interval_steps <- as.data.frame(steps_data[,j=list(average_steps=mean(steps, na.rm=TRUE)),by=interval])
plot(interval_steps$interval, interval_steps$average_steps, type = "l", xlab = "Interval", ylab = "Average Steps")
```

Find out which period has the highest average step count.

```{r}
interval_steps[which.max(interval_steps$average_steps),]$interval
```

8:35am has the highest average step count.

## Imputing missing values

```{r nas}
na_count <- sum(is.na(steps_data$steps))
```

The data set has `r na_count` NA values.

As a very basic first try, we shall create a copy of the dataset and replace all NAs with the average step count taken over all intervals measured.

```{r warning = FALSE}
int_ave <- mean(steps_data$steps, na.rm = TRUE)
steps_data_imput <- steps_data
steps_data_imput[is.na(steps_data_imput$steps)] <- int_ave
```

Now plot a histogram of the new data set.

```{r}
agg_steps_imput <- as.data.frame(steps_data_imput[,j=list(total_steps=sum(steps)),by=date])
hist(agg_steps_imput$total_steps, xlab = "Total Steps (NAs replaced)", main = "Histogram of Daily Total Steps (NAs replaced)")
```

Calculate the new mean and median, then compare with their previous values.

```{r new averages}
new_mn <- round(mean(agg_steps_imput$total_steps, na.rm=TRUE))
new_md <- median(agg_steps_imput$total_steps, na.rm=TRUE)
```

Imputing values in this fashion changes the mean from `r mn` to `r new_mn` and the median from `r md` to `r new_md`.
Both statistics show reasonable increases after NA values have been replaced however no judgement is made here as to whether the change is statistically significant or the methodology sound.

## Are there differences in activity patterns between weekdays and weekends?

Add a column to the imputed dataframe to distinguish between weekdays and weekends.

```{r}
steps_data_imput$dayname <- weekdays(as.POSIXlt(steps_data$date))
```

Create two subsets of the data, one for weekdays and another for weekends.

```{r}
weekdays <- subset(steps_data_imput, dayname %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
weekends <- subset(steps_data_imput, dayname %in% c("Saturday", "Sunday"))
```

Calculate average number of steps per five minute inteval for weekdays and weekends, then plot both datasets.

```{r}
wd_steps <- as.data.frame(weekdays[,j=list(average_steps=mean(steps, na.rm=TRUE)),by=interval])
we_steps <- as.data.frame(weekends[,j=list(average_steps=mean(steps, na.rm=TRUE)),by=interval])

par(mfrow = c(2,1))

plot(wd_steps$interval, wd_steps$average_steps, type = "l", xlab = "Interval", ylab = "Average Steps", main = "Weekdays")
plot(we_steps$interval, we_steps$average_steps, type = "l", xlab = "Interval", ylab = "Average Steps", main = "Weekends")
```

Weekend activity is sustained over the whole day, whereas weekday avtivity is more concentrated in the morning period, pre 10am.